---
title: "prenoms_dplyr_clean_data_if-at-all_en"
output:
  html_document: default
  pdf_document: default
---



Please, submit your work in shape of a zipped folder containing an RStudio project to xhec2018@thinkr.fr 

Deadline : 4th november, 11:00 PM 

The name of the project should be “dplyr_firstname_lastname”.

The project must contain a .Rmd file and the corresponding rendered .html file. Packages should be loaded at the top of the .Rmd file. 


```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Warm-up

### Make sure you've installed {dplyr} >= 0.7 and {prenoms} package

```{r eval=FALSE, echo=FALSE}


install.packages("operators")

```


### Load here `{dplyr}`, `{prenoms}` and any other needed package

```{r}

library(dplyr)
library(prenoms)
library(tidyr)
library(readr)
library(readxl)

```

### Import

#### prenomsdataset

Using `data(prenoms)` load `prenoms` dataset from  `{prenoms}` package.

```{r}
data("prenoms")

```

What kind of object is `prenoms` ? 

```{r}
class(prenoms)

```

Explore the database using the '5-functions-to-always-run-on-a-database'

```{r, eval=FALSE}


summarise(prenoms)
dim(prenoms)
head(prenoms)
View(prenoms)
colnames(prenoms)


```

Using `glimpse`, have a look at `prenoms`'s structure.

```{r}
glimpse(prenoms)

```

#### Regions, departements and surfaces

Load the "dpt_data_modif.csv" dataset from IGN (French public state administrative establishment founded in 1940[1] to produce and maintain geographical information for France and its overseas departments and territories) using the appropriate function. Data have been prepared for you: the surface of departement has been calculated and spatial data removed.

```{r}
depp_dataset <- read_csv("data/dpt_data_modif.csv")
```


#### Elementary and college schools

We also fetched for you on [data.gouv.fr](https://www.data.gouv.fr/fr/datasets/adresse-et-geolocalisation-des-etablissements-denseignement-du-premier-et-second-degres/#_) the addresses of "primary and secondary schools, the administrative structures of the Ministry of National Education. Public and private sectors."

1. Data preprocessing 
    + Import the csv file : "DEPP-etab-1D2D.csv" and name it "depp_orig"
        + Encoding is `"latin1"`
    + Transform zip code ("code_postal_uai") into 5 characters with zeros
    + Extract department numbers ("dpt") starting from column "code_postal_uai"
    + Save the modifications into "depp_modif.csv"

```{r}

#import data
depp_orig <- read_csv2("data/DEPP-etab-1D2D.csv")

#transform zip code into 5 characters with zeros
modif <- depp_orig
modif <- modif %>% mutate(code_postal_uai = case_when(
  nchar(code_postal_uai)<5 ~ paste("0", as.character(code_postal_uai), sep = ""),
  nchar(code_postal_uai)==5 ~ as.character(code_postal_uai)
)) %>% mutate(dpt = substr(code_postal_uai,1,2))


#Save modficiations
write_csv(modif, "data/depp_modif2.csv")
#new file created 

#modif %>% filter(is.na(dpt))
#depp_orig %>% filter(is.na(code_postal_uai))
```

2. Read the pre-processed "depp_modif.csv" file

```{r}
depp_modif <- read_csv("data/depp_modif.csv")
```


#### Facts observed by the police services and national gendarmerie units by department

We also gathered data from [data.gouv.fr](https://www.data.gouv.fr/fr/datasets/faits-constates-par-index-par-departement-et-par-annee/#_) concerning "all the facts observed by the police services and national gendarmerie units by department from 1996 to 2011"

1. Data preprocessing 
    - Import Excel sheet "2010" from "faitsconstatespardepartementde2002-a-2011.xls" file
        + _beware of the original formatting_
    - Copy it into "faits_2010_modif" in order to make some modifications:
        + Delete Excel calculations:
            + `Tout_département`, `Tout_index`
        + Transform in long format using `gather`
            + 4 columns : Index, Libellé, dpt, nombre
        + save the dataframe into a csv file "faits_2010_modif.csv"

```{r}
#read the file 
temp <- read_excel("data/faitsconstatespardepartementde2002-a-2011.xls", sheet="2010")

#Remove the column "Tout departements" and the row "Tout index"
temp2 <- temp %>%  
  select(colnames(temp)[-3]) %>%   
  filter(Départements != "Tout index")
 
#Change name of columns with departments number
colnames(temp2) <- temp2[1,]

#remove the department row
temp2 <- temp2 %>% 
  filter(Index != "Index")

#gather department and number into two distinct columns
temp2 <- temp2  %>% 
  gather(key=dpt, value=nombre, '01':'95') %>% 
  select(Index, Libellé, dpt, nombre)

#save the data frame into a csv file "faits_2010_modif.csv"
write_csv(temp2, "data/faits_2010_modif.csv")

```

2. Read preprocessed file "faits_2010_modif.csv"

```{r}
faits_2010 <- read_csv("data/faits_2010_modif.csv")

```


## Analyses

Some assumptions to do the exercise:

- every child born in a department stays into that department until the end of college
- every children between 11 and 14 years old is in a college
- the number of college is constant between 2010 and 2016
- College "à ouvrir" (i.e. "to be open") do not have children. Others have.

### Filter datasets to Metropolitan France

Datasets to be filtered: `prenoms`, `depp_modif`, `faits_2010_modif`, `dpt_data_modif`

- Department named "2A" and "2B" should be merged to "20"
- We only work with data in Metropolitan France, which means for "dpt" between `01` and `95` included. Others needs to be filtered.

```{r}


prenoms <- prenoms %>% 
  #replace all 2As and 2Bs occurence by 20
  mutate(dpt = replace(dpt,dpt %in% c("2A","2B"),"20")) %>% 
  #group by to merge lines of dpt 20 that are similar
  group_by(year,sex,dpt,name) %>%  
  #get the sum 
  summarise(n=sum(n)) %>%
  #group by year and dpt to find proportion for dpt 20
  group_by(year,dpt) %>% 
  #get proportion
  mutate(prop= n/sum(n)) %>% 
  #create a new column of dpt as double
  mutate(dpt_temp = as.double(dpt)) %>% 
  #filter dpt 01 to 95
  filter(dpt_temp >= 1 & dpt_temp <= 95) %>% 
  #remove the temp column
  select(-dpt_temp) %>% 
  #sort by dpt
  arrange(dpt) %>% 
  ungroup()


#same as before  
faits_2010 <- faits_2010 %>% 
  mutate(dpt = replace(dpt,dpt %in% c("2A","2B"),"20")) %>%
  group_by(Index,Libellé,dpt) %>%  
  summarise(nombre=sum(nombre)) %>% 
  arrange(dpt) %>% 
  ungroup()


#just remove dpt beyong 95
depp_modif <- depp_modif %>%
  mutate(dpt_temp = as.double(dpt)) %>% 
  filter(dpt_temp >= 1 & dpt_temp <= 95) %>% 
  select(-dpt_temp) %>% 
  arrange(dpt) %>% 
  ungroup()

#just merge 2A and 2B 
depp_dataset <- depp_dataset %>% 
  mutate(CODE_DEPT = replace(CODE_DEPT,CODE_DEPT %in% c("2A","2B"),"20")) %>%
  #here we remove the columns NOM_DEPT, CODE_CHF, NOM_CHF because they differ for 2A and 2B and we want   merge them 
  group_by(CODE_DEPT, CODE_REG, NOM_REG) %>%  
  summarise(surface_m = sum(surface_m)) %>% 
  ungroup()

```


### National average number of children per college in 2010 ?

```{r}

nb_college <- depp_modif %>% 
  filter(denomination_principale == "COLLEGE") %>% 
  summarise(count = n()) 


total <- prenoms %>% 
  filter(year >= 2010 - 14 & year <= 2010 - 11) %>% 
  summarise(total = sum(n)) 

propotion <- total/nb_college
propotion
```

### Average number of children per college in 2010 in each department?

- Arrange departments according to the calculated average in descending order

```{r}

#get number of children in college in 2010 per department
nb_students_by_dpt <- prenoms %>% 
 filter(year >= 2010 - 14 & year <= 2010 - 11) %>% 
  group_by(dpt) %>% 
  summarise(total = sum(n))

#joint nb_student_by_dpt and depp_modif to get a column with number of children in college in 2010 per department and number of colleges per department to get the average number
depp_modif %>% 
  group_by(dpt) %>% 
  summarise(nb_colleges_by_dpt = n()) %>% 
  inner_join(nb_students_by_dpt, by = "dpt") %>% 
  mutate(prop = total/nb_colleges_by_dpt) %>% 
  select(dpt, prop) %>% 
  arrange(desc(prop))


```

### Number of Facts observed by the police services in 2010 per department ? 

```{r}

facts_by_dpt <- faits_2010 %>% 
  group_by(dpt) %>% 
  summarise(nb_facts = sum(nombre))
  
facts_by_dpt


```

### Number of children born, number of colleges and facts related by the police services per department in 2010 ?

- Group all information in the same table
- Arrange by descending order of children, schools and facts

```{r}

table1 <- 
  #1)get number of children born in 2010
  prenoms %>% 
  filter(year == 2010) %>% 
  group_by(dpt) %>% 
  summarise(nb_children = sum(n)) %>% 
  #2)get number of colleges in 2010
  inner_join(
    depp_modif %>% 
    group_by(dpt) %>% 
    summarise(nb_colleges = n()),
    by = "dpt") %>% 
  #3)get number of facts in 2010
  inner_join(
    facts_by_dpt, 
    by = "dpt")

table1
```

### Number of children born, number of colleges and facts related by the police services per km² in 2010 by department?

```{r}
#convert surface from meters^2 to km^2
surface_by_dpt <- depp_dataset %>% 
  rename(dpt = CODE_DEPT, surface = surface_m) %>% 
  select(dpt, surface) %>% 
  mutate(surface = surface/10^6)

#add a column with the surface/dpt to table1 to get all the ratios we want
per_km <- table1 %>% 
  inner_join(surface_by_dpt, by = "dpt") %>% 
  mutate_at(vars(nb_children:nb_facts), funs(./surface)) %>% 
  rename(nb_children_p_km = nb_children, nb_colleges_p_km = nb_colleges, facts_p_km = nb_facts) %>% 
  select(-surface)

per_km
```

### Is there a correlation between the number of birth and the number of facts related by the police per km² in 2010 ?

```{r}
#correlation between the number of birth and the number of facts related by the police per km² in 2010 
cor(per_km$nb_children_p_km,per_km$facts_p_km, use = "complete.obs")

#plot of the relation
per_km %>% ggplot() + aes(x = nb_children_p_km, y = facts_p_km) +
  geom_point()
```

### What is the mean regional density (in number/km²) of the 15 most given first names in France ?

- Filter the 15 most given first names in France
- Create a wide table with the department as observations and the 15 most given names in columns (as variables): the count is at the row-column intersection 
- merge with the surface department infos
- Compute the region surface and the sum of names by region

```{r}

#compute 15 most given names in France in 2010
most_given_names <- 
  prenoms %>% 
  group_by(name) %>% 
  summarise(total = sum(n)) %>% 
  top_n(15, total) %>% 
  select(name) 
  

#compute the surface for each region
surface_by_region <- 
  depp_dataset %>% 
  group_by(NOM_REG) %>% 
  summarise(surface_reg = sum(surface_m))


temp <- 
  prenoms %>% 
  group_by(name, dpt) %>% 
  summarise(total = sum(n)) %>% 
  inner_join(most_given_names, by = "name") %>% 
  #get the 15 most given names in columns
  spread(key = name, value = total) %>%
  inner_join(
    depp_dataset %>% 
      rename(dpt = CODE_DEPT, surface_dpt = surface_m) %>% 
      select(dpt, surface_dpt, NOM_REG), 
    by = "dpt") %>% 
  inner_join(surface_by_region,
    by = "NOM_REG")


#replace all the NA in the column names by 0
temp[,2:16][is.na(temp[2:16])] <- 0

#add the new columns
temp %>%   
  mutate(sum_popular = rowSums(select(.,2:16))) %>% 
  mutate_at(vars(surface_reg, surface_dpt), funs(./10^6)) %>% 
  mutate(mean_regional_density = sum_popular/surface_reg)


```

#### Bonus question : map the mean regional density (in number/km²) of the 15 most given first names in France 

- Use the "department" shapefile to cross information and map data

```{r}





















```

